Tree establishment in models of forest dynamics: a key priority for further research

### 2023-01-24 manuscript/manuscriptRW.Rmd
added changes based on the comments in abstract and beginning of introduction

Tree establishment is a key process in forest dynamics, particularly in
the context of forest resilience after disturbance and climate change.
Models of forest dynamics have been in use since more than 50 years, but
an assessment of their capacity to represent tree establishment is
lacking. We assess how well current models are capturing the overall
abundance, species composition and mortality of tree establishment.

Using 15 models and model variants built to capture long-term (i.e.,
decadal to centennial) forest dynamics at the stand, landscape and
global levels, we simulate tree establishment at 200 sites representing
environmental gradients across Europe. The results are benchmarked
against comprehensive data from forest reserves.

Most of the models overestimate ingrowth rates, but this is not always
compensated by high simulated mortality rates in the early stages of
population dynamics. In the models, the effect of light availability on
ingrowth rates is captured better than the effect of temperature and
soil moisture, but the patterns are not consistent across models.
Simulated species diversity matches the observed ranges quite well.
Models simulating higher species diversity at the stand level do not
feature higher ingrowth diversity. The ingrowth of *Abies alba* is
captured across temperature and soil moisture gradients, whereas *Fagus
sylvatica* is generally underestimated. *Picea abies* ingrowth is
generally overestimated, particularly at colder sites. Increasing
complexity in the tree establishment modules of the models is not
related to higher accuracy of simulated tree establishment. Furthermore,
individual model design is more important than scale (stand, landscape,
global) and approach (empirical, process-based) to accurately capture
tree establishment.

Even though the overestimation of tree establishment represent a
challenge for the usefulness of model projections under uncertain
conditions such a climate change, and although some formulations should
be reconsidered to correct excessive ingrowth levels, it is remarkable
that most of the models capture the essential features of the highly
complex process of tree establishment, while not having been
parameterized with such a data. We conclude that in spite of all the
uncertainties, many models hold promise for assessing long-term forest
dynamics under changing environmental conditions.

@@

Forests provide a wide range of products and services of vital
importance to humankind. Under the pressure of climate change and
increasing disturbances, it is becoming ever more important to
understand how forest structure, composition and function will change,
and to evaluate forest capacity to adapt or be resilient after
disturbance [@lindner2010]. A wide range of models of forest dynamics
were developed over the past decades considering the impacts of climate
on forest dynamics [@bugmann2022]. From these studies, it is evident
that while we have a reasonably good understanding how to model tree
growth [@vanclay1997; @bugmann1996], and substantial efforts have been
dedicated to improving the depiction of tree mortality [@bugmann2019],
tree establishment remains a much less studied process that is often
represented rather coarsely in these models, which is a clear research
gap in the context of climate-induced forest disturbances and forest
resilience.

>>>>

Tree establishment is a key process in forest dynamics, particularly in
the context of forest resilience and climate change. Models of forest
dynamics have been in use for more than 50 years, but an assessment of
their capacity to represent tree establishment is lacking. We assess how
well current models capture the overall abundance, species composition
and mortality of tree establishment.

Using 15 models and model variants built to capture long-term (i.e.,
decadal to centennial) forest dynamics at the stand, landscape and
global levels, we simulate tree establishment at 200 sites representing
environmental gradients across Europe. The results are benchmarked
against comprehensive data from unmanaged forests.

Most of the models overestimate ingrowth rates, but this is not always
compensated by high simulated mortality rates in the early stages of
population dynamics. The effect of light availability on ingrowth rates
is captured better than the effect of temperature and soil moisture, but
patterns are not consistent across models. Simulated species diversity
of ingrowth matches the observed ranges quite well. Models simulating
higher species diversity at the stand level do not feature higher
ingrowth diversity. The ingrowth of *Abies alba* is captured across
temperature and soil moisture gradients, whereas ingrowth of *Fagus
sylvatica* is generally underestimated. *Picea abies* ingrowth is
generally overestimated, particularly at colder sites. Increasing
complexity in the tree establishment modules of the models is not
related to higher accuracy of simulated tree establishment. Furthermore,
individual model design is more important than scale (stand, landscape,
global) and approach (empirical, process-based) to accurately capture
tree establishment.

Even though the overestimation of tree establishment represents a
challenge for the usefulness of model projections under uncertain
conditions such as climate change, and although some formulations should
be reconsidered to correct excessive ingrowth levels, it is remarkable
that most of the models capture the essential features of the highly
complex process of tree establishment, while not having been
parameterized with such data. We conclude that in spite of all the
uncertainties, a lot can be gained by a focus on the modeling of
establishment processes in these model, this will make their long-term
forest dynamics projections under changing environmental conditions much
more robust.

@@

Forests provide a wide range of products and services of vital
importance to humankind [@foresteurope2020]. Under the pressure of
climate change and increasing disturbances, it is becoming ever more
important to understand how forest structure, composition and function
will change, and to evaluate forest capacity to adapt or be resilient
after disturbance [@lindner2010; @seidl2022]. A wide range of models of
forest dynamics were developed over the past decades considering the
impacts of climate on forest dynamics [@bugmann2022]. From these
studies, it is evident that while we have a reasonably good
understanding how to model tree growth [@vanclay1997; @bugmann1996], and
substantial efforts have been dedicated to improving the depiction of
tree mortality [@bugmann2019]. However, tree establishment remains a
much less studied process that is often represented rather coarsely in
these models [@walck2011; @leishman1992], which is a clear research gap
in the context of climate-induced forest disturbances and forest
resilience.
